---
title: "How does life expectancy change throughout the years in England"
author: "Daniel Jorden"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

INCLUDE LITERATURE REVIEW
REMOVE INSTALL.PACKAGES FROM CODE,JUST HAVE IT AS TEXT
If person doesn't have this packages, then install


```{r}

####How does life expectancy change across the years within England?

#Install relevant packages
install.packages("tidyverse")
install.packages("here")
install.packages("ggplot2")
install.packages("gganimate")  #Remove this from RMarkdown

#Load packages
library(tidyverse)
library(here)


#Loading in the data
data <- read.csv(here("data","trends in life expectancy.csv"))


```

##Data preperation
```{r}
#Remove first 5 rows
data <- data[-(1:5), ]

#Sanity check: check data starts from row 6
head(data)

#Remove header row and set first row as column names
colnames(data) <- data[1, ]
data <- data[-1, ]
colnames(data) <- make.names(colnames(data))


#Remove data relating to HCI and LCI, so that only life expectancy values remain
data_cleaned <- data %>%
  select(-contains("LCI"), -contains("UCI"))

#Change names in the Year column
colnames(data_cleaned) <- gsub("^X", "", colnames(data_cleaned)) # Remove the "X" at the start
colnames(data_cleaned) <- gsub("\\.to\\.", "-", colnames(data_cleaned)) #Replace ".to." with "-"
colnames(data_cleaned) <- gsub("\\.HLE$", "", colnames(data_cleaned)) #Remove ".HLE" at the end

#Sanity check: check data has first row as column names, displays the correct names, and only contains data for life expectancy for the regions I'm looking at
View(data_cleaned)


#Reshape dataset to long format
data_long <- data_cleaned %>% pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "Life.Expectancy")

#Check data is life expectancy column is numeric and convert if necessary
sapply(data_long, is.numeric)
data_long <- data_long %>% mutate(Life.Expectancy = as.numeric(Life.Expectancy))
is.numeric(data_long$Life.Expectancy)

#Calculate mean across gender for each region and year and assign it as the final prepared version of data
data_final <- data_long %>% group_by(Area.name, Year) %>%
  summarize(mean_life_expectancy = mean(Life.Expectancy, na.rm = TRUE)) %>% 
  filter(mean_life_expectancy != "NaN")

#Sanity check: Check data contains correct information and rows/columns are correct
View(data_final)




```

##Creating the visualisation

```{r}
#Load necessary packages
library(ggplot2)
library(gganimate)

#Define the mappings on the graph
p <- ggplot(data_final, aes(x = Year, y = mean_life_expectancy, color = Area.name, group = Area.name))

#Establishing the visualisation as a line graph: adding the lines for each region and the titles and key
p <- p + geom_line(linewidth = 1.2) + labs(title = "Trends in Life expectancy of regions in the UK across years", x = "Year", y = "Life Expectancy", color = "Region") + theme_minimal()

#Sanity check: check basic graph design is correct
p

```


##Improving the aesthetics of the graph
```{r}
#Add small points to highlight individual data values and rotate x-axis labels
p <- p + geom_point(size = 2) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Make Y axis longer
p <- p + scale_y_continuous(limits = c(0, 100))

```


Changing the scale on the Y-axis to 0-100 reduced the clarity of the graph as the lines were then too close together. I tried a range of different scales and decided to settle with 55-70 as this gave the clearest representation. I also changed the intervals at which the axis increases by
```{r}
#Changing Y axis scale and axis intervals
p <- p + scale_y_continuous(limits = c(55, 70))




```



